---
title: "HW 3"
output:
  word_document:
    fig_height: 5
    fig_width: 8
date: "Due 2/7/2025"
author: "No names - just add upon submission in gradescope"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)
options(show.signif.stars = FALSE)

library(ggplot2)
library(ggthemes)
library(tidyverse)
library(car)
library(effects)
library(mosaic)
library(remotes) #Needed for next two installs from github:

#Re-comment this after running it once:
#remotes::install_github("greenwood-stat/ggResidpanel")
library(ggResidpanel)
#remotes::install_github("greenwood-stat/catstats2")
library(catstats2)
library(gtsummary)
library(lubridate)
theme_set(theme_bw()) #Prevents need for + theme_bw() in ggplots
```

# HW 3 Instructions:

* In groups of 2 to 4 (your choice of group members, but you must be in a group to complete the submission), complete the following questions. If you are having trouble identifying group members, post in the Water Cooler channel in MS Teams.

Read Fabusova et al. (2024) and use the provided data set ("moths.xlsx") to answer the following questions.

* Fabusova M, Gaston KJ, Troscianko J. 2024 Pulsed artificial light at night alters moth flight behaviour. Biol. Lett. 20: 20240403. https://doi.org/10.1098/rsbl.2024.0403


The `Latency` (seconds) is the time to the moth starting to move after the treatment is applied or 10 seconds of dark for the control group. The treatment information is coded in three variables, `Treatment` with 5 levels (numbered unhelpfully as 1 to 5, shown with better labels once we figure out their meaning on the x-axis in Figure 2a) and also in `Colour` (cold or warm) and `Spectra` (LED or RGB - note that the LED level in this variable is the "Phosphor LED" and RGB level is the "RGB LED" in their Figure 2). 

1) The following code reads in the data set from their published Excel file and removes the `Sunset_time` that we will not use. What is the sample size of moths before any cleaning? Then use `drop_na` to clean the data set to remove any missing observations across the following variables: Date, ID, Location, Treatment, Treatment_y_n, Colour, Spectra, Moon_Phase, Temp, Wind_speed_ms, Humidity, Common_name, Latin_name, Family, Sub_family, Latency, Erratic_behaviour. What is the sample size after that cleaning? 

* Hint: The list above is provided so you can directly copy it into `drop_na`.

  **Sample size before cleaning is 393 observations. Sample size after cleaning is 302 observations.**

```{r}
library(readxl)
moth <- read_excel("moths.xlsx", 
     sheet = "complete_dataset_MAIN FAMILIES.", 
    na = "NA")

library(lubridate)
moth <- moth %>% 
  mutate(Date = factor(ymd(Date))) %>% 
  dplyr::select(-Sunset_time)

dat <- moth %>% drop_na(Date, ID, Location, Treatment, Treatment_y_n, Colour, Spectra, Moon_Phase, Temp, Wind_speed_ms, Humidity, Common_name, Latin_name, Family, Sub_family, Latency, Erratic_behaviour)
```


2) Read their Methods section (mainly parts A and B). The `Treatment`, `Colour`, and `Spectra` variables tell an interesting story about the reason for some of the missingness in the original data set. Use `tally(Spectra ~ Colour | Treatment, data = moth)` to explore those variables and select the correct SINGLE response from these choices:

a) The `Colour` and `Spectra` variables had missing values entered in order to remove outliers in the latency responses.

b) The `Colour` and `Spectra` variables were missing for the level 5 of `Treatment` because the researchers failed to record what treatment level was applied to the moths in those rows.

**c) The value 5 for the `Treatment` variable is a control group of moths that were not subjected to any light treatments, so they entered a missing value for the two levels of `Colour` and `Spectra`.**

**Answer is bolded**

```{r}
tally(Spectra ~ Colour | Treatment, data = dat)
```

3) Improve the factor levels of the `Treatment` variable in your cleaned data set to match the meaning of the levels using `fct_recode` and also create a natural log-transformed version of the `Latency` variable for later use. No discussion, just code here.

```{r}
dat <- dat %>% mutate(Treatment = factor(Treatment),
                      Treatment = fct_recode(Treatment,
                                             "Cool Phosphor-coated" = "1",
                                             "Warm Phosphor-coated" = "2",
                                             "Cool RGB" = "3",
                                             "Warm RGB" = "4"),
                      log_Latency = log(Latency))
```


## Use the cleaned data set for all remaining questions on the HW:

4) Using your improved labels for the `Treatment` variable, make a display of the `Latency` versus that grouping variable and then the logLatency versus that grouping variable. The second one should resemble their Figure 2b but be much more informative. What information does your _enhanced_ (hint!) display provide that is not presented in their version of the same plot?

  **The _enhanced_ stripchart displays the individual observations which allows us to see clustering of points and possible data trends that we may not be able to see in the boxplot.**

```{r}
enhanced_stripchart(Latency ~ Treatment, data = dat) + 
  labs(y = "Latency(seconds)")

enhanced_stripchart(log_Latency ~ Treatment, data = dat) + 
  labs(y = "Latency(log seconds)",
       title = "Initial Response Latency")
```

5) Make a scatterplot of the logLatency versus the Humidity faceted by Treatment with linear and nonparametric smoothing lines. Discuss the potential for an interaction between Humidity and Treatment on log-Latency based on the plot.

  **The smoothing lines are not the same across all treatment groups, suggesting a possible interaction between Humidity and Treatment on log-latency. This is especially noticible in the cool light sources.**

```{r}
dat %>% ggplot(mapping = aes(x = Humidity, y = log_Latency)) + 
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "lm") + 
  geom_smooth(col = "red", lty = 2, se = F) +
  facet_wrap(~ Treatment) + 
  scale_color_viridis(discrete = TRUE) +
  labs(x = "Humidity", 
       y = "Latency (log seconds)",
       title = "Plot of Latency (log secs) versus Humidity, Faceted by Treatment")

```


6) Fit the interaction model with Humidity and Treatment for log-latency. Write out the estimated model. Make sure you define all your indicator variables and use the notation from this course.

```{r}
moth_model <- lm(log_Latency ~ Humidity * Treatment, data = dat)

summary(moth_model)
```


$\hat{\mu}\{logLatency|Humidity \ast Treatment\} = -5.6 + 0.16Humidity + 5.04I_{Treatment=WarmPhosphor} + 11.5I_{Treatment=CoolRGB} + 1.5I_{Treatment=WarmRGB} - 0.099Humidity*I_{Treatment=WarmPhosphor} - 0.24Humidity*I_{Treatment=CoolRGB} - 0.015Humidity*I_{Treatment=WarmRGB}$

where $I_{Treatment=WarmPhosphor}$ is 1 when treatment is Warm phosphor-coated LED and 0 if else, 
$I_{Treatment=CoolRGB}$ is 1 when treatment is Cool RGB LED and 0 if else,
$I_{Treatment=warmRGB}$ is 1 when treatment is Warm RGB LED and 0 if else.  



7) The following work provides the simplified estimated models for two levels of the Treatment variable:

* Cold phosphor: $\hat{\mu}\{logLatency|Humidity \ast Treatment\} = -5.6 + 0.16Humidity$

* Cold RGB: $\hat{\mu}\{logLatency|Humidity \ast Treatment\} = 5.9 - 0.078Humidity$

Which of the following is the correct "size" interpretation (CIs are a topic for later - leave them as ellipses) for the 0.16 estimated slope coefficient for Humidity, assuming that this was the relative humidity they were using which is a percentage. Pick ONE - no discussion.

a) A one percent increase in relative humidity causes the estimated mean log-latency time to change by 0.16 log-seconds, controlled for treatment (95% CI of ... to ...).

b) For two otherwise similar moths but that had observed times that differed in relative humidity by 1 percent, the one measured during higher relative humidity had a higher estimated mean log-latency time by 0.16 log-seconds than the less humid observation, controlled for treatment (95% CI of ... to ...).

c) A one percent increase in relative humidity causes the estimated mean log-latency time to change by 0.16 log-seconds for an observation with a cold phosphorous light (95% CI of ... to ...).

**d) For two otherwise similar moths but that had observed times that differed in relative humidity by 1 percent, the one measured during higher relative humidity had a higher estimated mean log-latency time by 0.16 log-seconds than the less humid observation for an observation with a cold phosphorous light (95% CI of ... to ...).**

**Answer is bolded**  

8) Drop the interaction from the model and re-fit it. Make an effects plot for the model. Write a size interpretation for the `humidity` slope coefficient from that model, making sure you find and include the 95% CI in parentheses in the size interpretation.

```{r fig.width = 12, fig.height = 8}
moth_model2 <- lm(log_Latency ~ Humidity + Treatment, data = dat)
summary(moth_model2)
confint(moth_model2, level = 0.95)

plot(allEffects(moth_model2))
```

$\hat{\mu}\{logLatency|Humidity + Treatment\} = -1.69 + 0.077Humidity + 0.63I_{Treatment=WarmPhosphor} + 0.65I_{Treatment=CoolRGB} + 0.84I_{Treatment=WarmRGB}$

where $I_{Treatment=WarmPhosphor}$ is 1 when treatment is Warm phosphor-coated LED and 0 if else, 
$I_{Treatment=CoolRGB}$ is 1 when treatment is Cool RGB LED and 0 if else,
$I_{Treatment=warmRGB}$ is 1 when treatment is Warm RGB LED and 0 if else.  

**For two otherwise similar moths but that had observed times that differed in relative humidity by 1 percent, the one measured during higher relative humidity had a higher estimated mean log-latency time by 0.08 log-seconds than the less humid observation, controlled for treatment (95% CI of -0.01 to 0.16).**


9) Document any collaborations or pertinent discussions with other students or resources outside of your group and the resources that I am providing and the _Sleuth_ that you used to complete this assignment *or report that you did not have any*. If you used generative AI (chatGPT, Bard, etc.), report which question(s) and how/what you asked for.

  * **NONE**